Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-12-09 index.qmd
section:issues; process:write

<!-- 2. Example of discriminal processes -->
@fig-discriminal_process illustrates the discriminal processes along a quality trait continuum for two written texts. The figure shows that these processes individually follow a Normal distribution. Furthermore, it shows that the modal discriminal process for Text B is positioned further along the continuum than Text A $(S_{B} > S_{A})$. Additionally, the figure highlights the broader distribution of Text B compared to Text A as a result of its larger discriminal dispersion $(\sigma_{B} > \sigma_{A})$.

@@

<!-- 4. Example of discriminal difference -->
@fig-discriminal_difference illustrates the distribution of the discriminal difference for the texts presented in @fig-discriminal_process. The figure indicates that the judge perceives Text B as having significantly higher quality than Text A. This conclusion is evident from the positive difference in the modal discriminal processes of the texts $(S_{B} - S_{A} > 0)$ and the area under the curve where the discriminal difference distinctly favors Text B over Text A, denoted as $P(B > A)$ (shaded gray area). As a result, the dichotomous outcome of this comparison is more likely to favor Text B over Text A.

@@

<!-- 1. Thurstone's Cases and their simplifying assumptions  -->
The previous section outlines the general form of Thurstone's theory, which applies to a CJ design where a single judge evaluates multiple stimuli. For the practical application of the theory, Thurstone developed four additional cases derived from this general form, where each successive case incorporates additional simplifying assumptions. Case I represents the general form of the theory. Case II extends this by allowing multiple judges to make comparisons rather than restricting the comparisons to a single judge. Case III introduces the assumption of zero correlation between stimuli. Case IV builds on this by assuming the stimuli have similar dispersions. Finally, Case V replaces this assumption with the condition that the stimuli have equal discriminal dispersions. @tbl-thurstone_cases summarizes these cases and their assumptions. For a detailed discussion of this progression, refer to @Thurstone_1927b and @Bramley_2008 [pp. 248-253].

@@

<!-- 3. But what are limitations of Case V (and the BTL model)? -->
Nevertheless, Thurstone originally developed Case V to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp. 269], prioritizing statistical simplicity over precision in trait measurement [@Kelly_et_al_2022, pp. 677]. He explicitly warned against its untested application, stating that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp. 270], acknowledging that some assumptions might be problematic when complex traits or heterogeneous stimuli are involved [@Thurstone_1927a, pp. 376]. Consequently, given that modern CJ applications frequently involve such traits and stimuli, two main assumptions of Case V and, by extension, of the BTL model may not consistently hold in theory or practice: the assumption of equal dispersion and zero correlation between stimuli.

@@

<!-- 1. What does the equal dispersion between stimuli imply? -->
The discriminal dispersions of stimuli are crucial in determining the dichotomous outcomes of pairwise comparisons. Discrepancies in these dispersions shape the distribution of the discriminal difference, directly influencing the comparison outcome. @fig-dispersion illustrates how more uncertainty in the trait perception of one text relative to another broadens the distribution of their discriminal difference. This broadening affects the area under the curve where the discriminal difference distinctly favors one text over the other, $P(B > A)$. Additionally, the figure shows that when the discriminal dispersions of the texts are equal $(\sigma_{B}-\sigma_{A}=0)$, the discriminal difference is more likely to favor one text over the other than when their dispersions differ (shaded gray area).

![The Discrepancy in the dispersions of stimuli and their effect on the distribution of the discriminal difference](/images/figures/dispersion.png){#fig-dispersion width=70%}

<!-- 2. But does it hold?  -->
However, Thurstone contended that the assumption of equal dispersions may not hold when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp. 376], as these traits and stimuli can introduce judgment discrepancies due to their unique features [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022]. Indeed, evidence of such violation may already exist in the CJ literature as misfit statistics. These statistics measure the judgment discrepancies associated with a given stimulus [@Pollitt_2004, pp. 12; @Goossens_et_al_2018, pp. 20]. For instance, labeling texts as "misfits" indicates that comparisons involving these texts result in more judgment discrepancies than others [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018]. This finding implies that discriminal differences associated with "misfits" texts usually display a broader dispersion, suggesting that their discriminal processes also exhibit more variation than other texts. Notably, this reasoning also applies to "misfit" judges, whose evaluations reflect substantial deviations from the shared consensus due to the unique characteristics of the stimuli or the judges themselves. Moreover, the presence of these "misfit" judges and their deviations can introduce additional statistical and measurement issues, as discussed in @sec-theory-issue1b.

@@

<!-- 3. But what are the challenges when these assumptions are violated? -->
Therefore, assuming equal dispersions between stimuli, despite its violation, can cause Case V (and the BTL model) to inflate the reliability of the comparison outcome. Moreover, ignoring the difference in dispersions can lead to the neglect of critical differences in the reliability of the trait across stimuli, resulting in inaccurate conclusions about the trait's estimates [@McElreath_2020, pp. 370]. Furthermore, if researchers acknowledge that misfit statistics help identify these critical differences, the usual practice in the CJ literature of excluding stimuli based on these statistics [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018] risks discarding valuable information and introducing bias into the trait's estimates [@Zimmerman_1994; @McElreath_2020, chap. 12]. The direction and magnitude of these biases are unpredictable, as they depend on the specific stimuli researchers exclude from the analysis. Together, these oversights undermine the reliability of the trait and ultimately compromise its validity [@Perron_et_al_2015, pp. 2].

@@

<!-- 1. What does the zero correlation between stimuli imply? -->
Similar to the discriminal dispersions, the correlation between discriminal processes plays a crucial role in determining the dichotomous outcomes of pairwise comparisons. This correlation also shapes the distribution of the discriminal difference, directly influencing the comparison outcome. @fig-correlation illustrates how the dependence in trait perception between two texts narrows the distribution of their discriminal difference. This narrowing affects the area under the curve where the discriminal difference distinctly favors one text over the other, $P(B>A)$. Moreover, the figure shows that when two texts are independent or uncorrelated $(\rho=0)$, their discriminal difference is less likely to favor one text over the other than when the texts are highly correlated (shaded gray area).

@@

<!-- 4. The second scenario where the assumption does not hold -->
In the second scenario, the shared context or inherent connections created by additional hierarchical structures may further introduce dependencies between stimuli, a statistical phenomenon commonly known as clustering [@Everitt_et_al_2010]. Although the CJ literature acknowledges the presence of such hierarchical structures, the statistical handling of this extra source of dependency between stimuli has been inadequate. For example, when CJ data includes multiple samples of stimuli from the same individuals, researchers often rely on (average) estimated BTL scores to conduct subsequent analyses and tests at the individual hierarchical level [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021]. This approach, however, can introduce additional statistical and measurement issues, as discussed in @sec-theory-issue2.

<!-- 5. But what are the challenges when this assumption is violated? -->
In any case, the psychometric and statistical literature emphasizes the need to address factors that create dependencies between stimuli, as failing to do so can affect the reliability of the comparison outcomes and lead to inaccurate conclusions about the trait's estimates. For instance, researchers who overlook additional relevant traits, such as judges' biases, can cause dimensional mismatches in the statistical model used for analysis. This mismatch can artificially inflate the reliability of the trait [@Hoyle_et_al_2023, pp. 341] or, even worse, introduce bias into the trait's estimates [@Ackerman_1989]. Furthermore, as discussed in @sec-theory-issue1a, researchers who exclude judges based on misfit statistics can risk discarding valuable information, further biasing the trait's estimates [@Zimmerman_1994; @McElreath_2020, chap. 12]. Lastly, researchers who fail to account for hierarchical (grouping) structures can reduce the precision of model parameter estimates, which may amplify the overestimation of reliability [@Hoyle_et_al_2023, pp. 482]. Together, these issues undermine the reliability of the trait and compromise the validity of its estimates [@Perron_et_al_2015, pp. 2].

@@

@@

@@

<!-- 3. What is the solution? -->

<!-- Notes:  -->
<!-- - There is some gain on considering a more integrated overarching systematic way of looking on what happen in DCJ when people compare two stimuli -->
<!-- - This paper tries to systematically integrate all aspects at play in a single theoretical model, we will build the model in a stepwise manner using -->
<!-- - we will build a scientific model (more theoretical model) that integrates different aspects of the method that are at play when people use DCJ. -->

>>>>

<!-- 2. Example of discriminal processes -->
@fig-discriminal_process illustrates the discriminal processes along a quality trait continuum for two written texts. The figure shows that each discriminal process follows a Normal distribution. It also indicates that the modal discriminal process for Text B is positioned further along the continuum than that of Text A $(S_{B} > S_{A})$, signifying a higher quality for Text B. Additionally, the figure highlights that Text B has a broader distribution compared to Text A, which arises from its larger discriminal dispersion $(\sigma_{B} > \sigma_{A})$.

<!-- 1. Thurstone's Theory of CJ: discriminal process -->
<!-- In its most general form, Thurstone’s theory addresses pairwise comparisons of stimuli made by a single judge [@Thurstone_1927b, pp. 267]. The theory posits that two key factors determine the dichotomous outcome of these comparisons: the discriminal process of each stimulus and their discriminal difference. The *discriminal process* represents the psychological impact each stimulus has on judges or, more simply, their perception of the stimulus trait. For instance, according to the theory, the discriminal process for a text along a quality trait continuum follows a Normal distribution, as depicted in @fig-discriminal_process. In this distribution, the mode or mean, termed the *modal discriminal process*, represents the stimulus's position on the trait continuum. @fig-discriminal_process show the modal discriminal process for Text B is positioned further along the continuum than that of Text A $(S_{B} > S_{A})$. Additionally, the dispersion of this distribution, called the *discriminal dispersion*, captures variability in the perceived trait of the stimulus. @fig-discriminal_process highlights that Text B has a broader distribution compared to Text A, which arises from its larger discriminal dispersion $(\sigma_{B} > \sigma_{A})$. -->

@@

<!-- 4. Example of discriminal difference -->
@fig-discriminal_difference illustrates the distribution of the discriminal difference for the texts presented in @fig-discriminal_process. The figure indicates that the judge perceives Text B as having significantly higher quality than Text A. This conclusion is evident from the positive difference in their modal discriminal processes $(S_{B} - S_{A} > 0)$ and the area under the curve where the discriminal difference distinctly favors Text B over Text A, denoted as $P(B > A)$ (shaded gray area). As a result, the dichotomous outcome of this comparison is more likely to favor Text B over Text A.

@@

<!-- 1. Thurstone's Cases and their simplifying assumptions  -->
Since the general form of Thurstone's theory, outlined in @sec-thurstone_theory, applies to a CJ design where a single judge evaluates multiple stimuli, Thurstone developed four additional cases for the theory's practical application, each progressively incorporating additional simplifying assumptions. Case I represents the general form of the theory. Case II extends this by allowing multiple judges to make comparisons rather than restricting the comparisons to a single judge. Case III introduces the assumption of zero correlation between stimuli. Case IV builds on this by assuming the stimuli have similar dispersions. Finally, Case V replaces this assumption with the condition that the stimuli have equal discriminal dispersions. @tbl-thurstone_cases summarizes these cases and their assumptions. For a detailed discussion of this progression, refer to @Thurstone_1927b and @Bramley_2008 [pp. 248-253].

@@

<!-- 3. But what are limitations of Case V (and the BTL model)? -->
Nevertheless, Thurstone originally developed Case V to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp. 269], prioritizing statistical simplicity over precision in trait measurement [@Kelly_et_al_2022, pp. 677]. Thurstone explicitly warned against its untested application, stating that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp. 270], acknowledging that some assumptions could be problematic when researchers asesss complex traits or heterogeneous stimuli [@Thurstone_1927a, pp. 376]. Consequently, given that modern CJ applications frequently involve such traits and stimuli, two main assumptions of Case V and, by extension, of the BTL model may not consistently hold in theory or practice: the assumption of equal dispersion and zero correlation between stimuli.

@@

<!-- 1. What does the equal dispersion between stimuli imply? -->
The discriminal dispersions of stimuli are crucial in determining the dichotomous outcomes of pairwise comparisons. Holding all other modeling factors constant, discrepancies in these dispersions shape the distribution of the discriminal difference, directly influencing the comparison outcome. @fig-dispersion illustrates how more uncertainty in the trait perception of one text relative to another, $(\sigma_{B}-\sigma_{A})$, broadens the distribution of their discriminal difference. This broadening affects the area under the curve where the discriminal difference distinctly favors one text over the other, $P(B > A)$. Additionally, the figure shows that when the discriminal dispersions of the texts are equal $(\sigma_{B}-\sigma_{A}=0)$, the discriminal difference is more likely to favor one text over the other than when their dispersions differ (shaded gray area).

![The discrepancy in the dispersions of stimuli and their effect on the distribution of the discriminal difference](/images/figures/dispersion.png){#fig-dispersion width=70%}

<!-- 2. But does it hold?  -->
However, Thurstone contended that this assumption of equal dispersions may not hold when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp. 376], as these traits and stimuli can introduce judgment discrepancies due to their unique features [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022]. Indeed, evidence of such violation may already exist in the CJ literature as misfit statistics. These statistics measure the judgment discrepancies associated with a given stimulus [@Pollitt_2004, pp. 12; @Goossens_et_al_2018, pp. 20]. For instance, labeling texts as "misfits" indicates that comparisons involving these texts result in more judgment discrepancies than others [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018]. This finding implies that discriminal differences associated with "misfits" texts usually display a broader dispersion, suggesting that the discriminal processes of these texts also exhibit more variation than other texts. Notably, this reasoning also applies to "misfit" judges, whose evaluations reflect substantial deviations from the shared consensus due to the unique characteristics of the stimuli or the judges themselves. Moreover, the presence of these "misfit" judges and their deviations can introduce additional statistical and measurement issues, which we discuss in @sec-theory-issue1b.

@@

<!-- 3. But what are the challenges when these assumptions are violated? -->
Therefore, assuming equal dispersions between stimuli, despite its violation, can cause Case V (and the BTL model) to inflate the reliability of the outcome, resulting in inaccurate conclusions about the comparison. Moreover, ignoring the disparity in dispersions can lead to the neglect of critical differences in the reliability of the trait across these stimuli, leading to erroneous conclusions about the trait's estimates [@McElreath_2020, pp. 370]. Additionally, if researchers acknowledge that misfit statistics help identify these critical differences in dispersion, the usual practice in the CJ literature of excluding stimuli based on these statistics [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018] risks discarding valuable information and introducing bias into the trait's estimates [@Zimmerman_1994; @McElreath_2020, chap. 12]. The direction and magnitude of these biases are often unpredictable, as they depend on the specific stimuli researchers exclude from the analysis.

@@

<!-- 1. What does the zero correlation between stimuli imply? -->
Similar to the discriminal dispersions, the correlation between discriminal processes plays a crucial role in determining the dichotomous outcomes of pairwise comparisons. Holding all other modeling factors constant, this correlation shapes the distribution of the discriminal difference, directly influencing the comparison outcome. @fig-correlation illustrates how the dependence in trait perception between two texts narrows the distribution of their discriminal difference. This narrowing affects the area under the curve where the discriminal difference distinctly favors one text over the other, $P(B>A)$. Moreover, the figure shows that when two texts are independent or uncorrelated $(\rho=0)$, their discriminal difference is less likely to favor one text over the other than when the texts are highly correlated (shaded gray area).

@@

<!-- 4. The second scenario where the assumption does not hold -->
In the second scenario, the shared context or inherent connections created by additional hierarchical structures may further introduce dependencies between stimuli, a statistical phenomenon commonly known as clustering [@Everitt_et_al_2010]. Although the CJ literature acknowledges the presence of such hierarchical structures, the statistical handling of this extra source of dependency between stimuli has been inadequate. For example, when CJ data includes multiple samples of stimuli from the same individuals, researchers often rely on (average) estimated BTL scores to conduct subsequent analyses and tests at the individual hierarchical level [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021]. However, this approach can introduce additional statistical and measurement issues, which we discuss in @sec-theory-issue2.

<!-- 5. But what are the challenges when this assumption is violated? -->
In any case, the psychometric and statistical literature emphasizes the need to address factors that create dependencies between stimuli, as failing to do so can affect the reliability of the comparison outcomes and lead to inaccurate conclusions about the trait's estimates. For instance, researchers who overlook additional relevant traits, such as judges' biases, can cause dimensional mismatches in the statistical model used for analysis. This mismatch can artificially inflate the reliability of the trait [@Hoyle_et_al_2023, pp. 341] or, even worse, introduce bias into the trait's estimates [@Ackerman_1989]. Furthermore, as discussed in @sec-theory-issue1a, researchers who exclude judges based on misfit statistics can risk discarding valuable information, further biasing the trait's estimates [@Zimmerman_1994; @McElreath_2020, chap. 12]. Lastly, researchers who fail to account for hierarchical (grouping) structures can reduce the precision of model parameter estimates, which may amplify the overestimation of the trait's reliability [@Hoyle_et_al_2023, pp. 482].

@@

In aggregate, researchers' inadequate handling of violations to the assumptions of equal dispersion and zero correlation between stimuli, along with the apparent disconnect between CJ's approach to trait measurement and hypothesis testing, can undermine the reliability of the trait and ultimately compromise its validity [@Perron_et_al_2015, pp. 2]. Consequently, adopting a more systematic and integrated approach to examining what happens when judges compare two stimuli could offer several statistical and measurement benefits, including addressing these issues.

@@

This section presents a theoretical model for CJ that extends Thurstone's theory. The model systematically incorporates all factors involved when judges make pairwise comparisons. Additionally, the section develops the statistical translation of the theoretical model based on assumptions informed by the CJ theory.

<!-- in a stepwise manner -->

@@
